---
title: "README"
author: "Julien"
date: "23 Jan 2015"
output: html_document
---

This file describes the script instructions necessary to build the required tidy dataset.

Downloading the data
====================

We use a classical `download.file` function with the URL provided on the Coursera website.
The zipped downloaded file is stored in a `data` folder within the working directory.

```{r,echo=FALSE,eval=FALSE}
if (!file.exists("./data")){dir.create("./data")}
fileURL <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
download.file(fileURL,destfile="./data/ds.zip",mod="wb",method="curl")
```

We then use the `unzip` function, which works according to its name. Within the working directory a new directory is created: `UCI HAR Dataset` with all the data in it.

```{r,echo=FALSE,eval=FALSE}
unzip("./data/ds.zip")
```

Then we read the `.txt` files using the `read.table` function on
* the features list: `features.txt` (subdirectory /UCI HAR Dataset): data frame `features`
* the training data: `X_train.txt`, `y_train.txt`, `subject_train.txt` (subdirectory /UCI HAR Dataset/train): data frames `xtrain`, `ytrain`, `subjecttrain`
* the test data: `X_test.txt`, `y_test.txt`, `subject_test.txt` (subdirectory /UCI HAR Dataset/test): data frams `xtest`, `ytest`, `subjecttest`

```{r}
features <- read.table("./UCI HAR Dataset/features.txt",header=FALSE)
features <- features[,2]
features <- as.character(features)

xtrain <- read.table("./UCI HAR Dataset/train/X_train.txt",header=FALSE)
ytrain <- read.table("./UCI HAR Dataset/train/y_train.txt",header=FALSE)
subjecttrain <- read.table("./UCI HAR Dataset/train/subject_train.txt",header=FALSE)
xtest <- read.table("./UCI HAR Dataset/test/X_test.txt",header=FALSE)
ytest <- read.table("./UCI HAR Dataset/test/y_test.txt",header=FALSE)
subjecttest <- read.table("./UCI HAR Dataset/test/subject_test.txt",header=FALSE)
```


Merging the data
================

We use the `rbind` function to merge training and test sets and we create three new data frames: `x`, `y`, `subject`.

```{r,echo=FALSE}
x <- rbind(xtrain,xtest)
y <- rbind(ytrain,ytest)
subject <- rbind(subjecttrain,subjecttest)
```


Extracting the measurements on the mean and the standard deviation for each measurement
=======================================================================================

We write a function `is_mean_or_std()` that detects, within the features names, whether a feature is either a mean or a standard deviation, by extracting strings. It returns `TRUE` is so.

```{r,echo=FALSE}
is_mean_or_std <- function(x){
    y <- substr(x,nchar(x)-8,nchar(x)-2)
    if (y == "-mean()"){return(TRUE)}
    y <- substr(x,nchar(x)-7,nchar(x)-2)
    if (y == "-std()"){return(TRUE)}
    y <- substr(x,nchar(x)-5,nchar(x))
    if (y == "-std()"){return(TRUE)}
    y <- substr(x,nchar(x)-6,nchar(x))
    if (y == "-mean()"){return(TRUE)}
    FALSE
}
```

Then we store the index of `features` which satisfies the `is_mean_or_std` criteria into a `extract_measurement` variable and select these features in `x`:

```{r,echo=FALSE}
extract_measurement <- which(sapply(features,is_mean_or_std))
x <- x[,extract_measurement]
```

There are 66 such variables.

Using descriptive activity names to name the activities in the dataset
=====================================================================

We read the activity list from the `activity_labels.txt` file.

```{r,echo=FALSE}
activities <- read.table("./UCI HAR Dataset/activity_labels.txt",header=FALSE)
activities <- activities[,2]
```

We use the `mutate` function in the `dplyr` package to match the `y` levels with the corresponding activities.

```{r,echo=FALSE}
library(dplyr)
y <- mutate(y,activity = activities[y[,1]])
```

And we merge `x` and `y$activity` and `subject` into `x`.

```{r,echo=FALSE}
x <- cbind(x,y$activity,subject)
```


Labelling the dataset with descriptive variable names
=====================================================

We simply set the names of the `x` dataset using the selected `features` (according to `extract_measurement`), as well as `activity` (variable coming from `y`) and `individual` (variable coming from `subject`).

```{r,echo=FALSE}
names(x) <- c(features[extract_measurement],"activity","individual")
```


Creating an independant, tidy dataset with the average of each variable for each activity and each subject
==========================================================================================================

We use the `dplyr` package to `group_by` both `activity` and `individual`. We compute the mean for each other variable using the `summarise_each` function. The data is stored into the `final` data frame.

```{r,echo=FALSE}
final <- x %>% group_by(activity,individual) %>% summarise_each(funs(mean))
```

Writing the results
===================

We use the `write.table` function to export the `final` data frame into the `final.txt` file.

```{r}
write.table(final,"./final.txt",row.name=FALSE)
```

